Longitudinal Principal Manifold Estimation

### 2024-06-14 paper/main.tex

Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This approach mitigates the effects of random disturbances to the manifold between time points. In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

\subsection{A Motivating Question from the Medical Imaging}

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain image of a cognitively normal person which compose the majority of the data used for training the segmentation models. In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

@@

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, including time in the dimension reduction process. Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time. This work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript. The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.

@@

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud (i.e., $D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest (i.e., $d=2$). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a function $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $r \to \left( \frac{\partial ^2 f}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\mathbf{X}$ has compact support and finite second moments. Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:

@@

&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\mathbf{X} - f(\pi_g(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f). \nonumber

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem. \cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form $$f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ l = 1, 2, \dots, D,$$
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud $\bf{X}$. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

@@

The PME framework described in Section \ref{s:PME} is an excellent approach for fitting a 2-dimensional surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. In this section, we introduce an approach to extending the PME framework to longitudinal clouds of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant. Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit. Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

This problem can be posed for positive integer values of $D$ and $d$ with $d < D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time.

@@

>>>>

Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This approach mitigates the effects of random disturbances to the manifold between time points. In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}

@@

\subsection{A Motivating Question from Medical Imaging}

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models. In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.

@@

Nonlinear manifold learning methods have been extensively studied in the literature. However, many of them are not applicable in longitudinal studies, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}). That is, these methods include the time dimension in the dimension reduction process, as a result of which the time dimension is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension. We use the oldest linear manifold learning method, principal component analysis (PCA, \cite{Pearson1901on}), as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. In addition, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation~\eqref{eq: spline representation of the PME}, in Section \ref{s:PME}). Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

%Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, including time in the dimension reduction process. Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

%Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis (PCA) method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

%Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

This work was motivated by longitudinal imaging studies, where data of participants are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this article. The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources, including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.

@@

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure \ref{fig:adni_result}). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments. Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:

@@

\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2. 

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem. \cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form
\begin{align}\label{eq: spline representation of the PME}
    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud. Importantly, $N$ is generally much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

@@

In this section, we introduce an approach to extending the PME framework described in Section \ref{s:PME} to longitudinal point clouds. We first heuristically explain the goal of our proposed model and its importance in the segmentation of brain regions for longitudinal imaging studies. Our goal is to estimate a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. The estimated surfaces across different time points enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant. Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit, and $(x_1, x_2, x_3)$ denotes a spatial point on the surface of the segmented brain region of interest.

Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data, in which 2-dimensional point clouds are observed longitudinally at several time points. The goal in this simplified example is to fit a surface in 3-dimensional spacetime---comprising two spatial dimensions and one time dimension---that fits the longitudinal data $(x_1, x_2, t_j)$ in spacetime and is not affected by the random noise added at each time point and in space (see Figure \ref{fig:lpme_step4}). The PME framework does not directly apply due to the following:
\begin{enumerate}
    \item Suppose we apply the PME approach to the data $(x_1, x_2, t_j)$ in spacetime and obtain an estimated surface $(r_1, r_2) \mapsto F(r_1, r_2)\in\mathbb{R}^3$ in spacetime. It is likely that neither coordinate $r_1$ nor $r_2$ coincides with the time dimension---they are (potentially nonlinear) combinations of both time and spatial coordinates, which prevents meaningful interpretation of the time dimension.

    \item If the manifold is estimated separately at each time point, the results may not be smooth over time.
\end{enumerate}
We expect that our proposed dimension reduction method will yield a smooth surface $(r,t)\mapsto F(r,t)$ in spacetime while ensuring that one coordinate, $t$, coincides with the time dimension. This can be achieved by the longitudinal principal manifold estimation (LPME) framework proposed in this section.

%The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

%Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.

@@

Hereafter, we utilize the following notations to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each individual $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by $(\mathbf{r},t)\mapsto F(\mathbf{r},t)=:f_t(\mathbf{r})\in\mathbb{R}^D$, where $\mathbf{r}\in\mathbb{R}^d$, and $t$ coincides with the time dimension.
